---
title: "Condensed Bishop"
author: "Aimee Barciauskas"
date: "21 October 2015"
output: rmarkdown::tufte_handout
---

# Introduction

These are my condensed notes from Bishop's **Pattern Recognition and Machine Learning[^Bishop's Pattern Recognition and Machine Learning].**

\newthought{**This is a work in progress**}

## Section 2.4 The Exponential Family

The exponential family of functions can be generalized as:

$p(\textbf{x}|\eta) = h(x) g(\eta) e^{\eta^{T} u(x) }$

&nbsp;

Where:

* Where $h(x)$ is ??
* And $g(\eta)$ is the coefficient that ensures the distribution is normalized.
* And $u(x)$ is som function of $x$

The generalized form of the likelihood estimator function for the exponential family of functions is:

&nbsp;

$p(\textbf{x}|\eta) = \prod_{n=1}^{N} h(x_{n}) g(\eta)^{N} e^{\eta^{T} \sum_{n=1}^{N}u(x_{n}) }$

____________________
# NOT MY STUFF BELOW

# Figures

## Margin Figures

Images and graphics play an integral role in Tufte's work. To place figures or tables in the margin you can use the `fig.margin` knitr chunk option. For example:

```{r, fig.margin = TRUE, fig.cap = "Sepal length vs. petal length, colored by species"}
library(ggplot2)
qplot(Sepal.Length, Petal.Length, data = iris, color = Species)
```

Note the use of the `fig.cap` chunk option to provide a figure caption. You can adjust the proportions of figures using the `fig.width` and `fig.height` chunk options. These are specified in inches, and will be automatically scaled down to fit within the handout margin.

## Equations

You can also include \LaTeX\ equations in the margin by explicitly invoking the `marginfigure` environment.

\begin{marginfigure}
$$\frac{d}{dx}\left( \int_{0}^{x} f(u)\,du\right)=f(x).$$
\caption{An equation}
\end{marginfigure}

Note the use of the `\caption` command to add additional text below the equation.

## Full Width Figures

You can arrange for figures to span across the entire page by using the `fig.fullwidth` chunk option.

```{r, fig.width = 10, fig.height = 2, fig.fullwidth = TRUE, fig.cap = "Full width figure"}
qplot(wt, mpg, data=mtcars, colour=factor(cyl))
```

Note the use of the `fig.width` and `fig.height` chunk options to establish the proportions of the figure. Full width figures look much better if their height is minimized.

## Main Column Figures

Besides margin and full width figures, you can of course also include figures constrained to the main column.

```{r, fig.cap = "Another figure"}
qplot(factor(cyl), mpg, data = mtcars, geom = "boxplot")
```

# Sidenotes

One of the most prominent and distinctive features of this style is the extensive use of sidenotes. There is a wide margin to provide ample room for sidenotes and small figures. Any use of a footnote will automatically be converted to a sidenote. ^[This is a sidenote that was entered using a footnote.]

If you'd like to place ancillary information in the margin without the sidenote mark (the superscript number), you can use the `\marginnote` command. \marginnote{This is a margin note.  Notice that there isn't a number preceding the note.}

Note also that the two footnote references (`tufte_latex` and `books_be`, both defined below) were also included in the margin on the first page of this document.

# Tables

You can use the **xtable** package to format \LaTeX\ tables that integrate well with the rest of the Tufte handout style. Note that it's important to set the `xtable.comment` and `xtable.booktabs` options as shown below to ensure the table is formatted correctly for inclusion in the document.

```{r, results='asis'}
library(xtable)
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(head(mtcars[,1:6]), caption = "First rows of mtcars")
```


[^Bishop's Pattern Recognition and Machine Learning]: http://www.rmki.kfki.hu/~banmi/elte/Bishop%20-%20Pattern%20Recognition%20and%20Machine%20Learning.pdf
