# Background Exercises GLM

## Mean and variance of binary data

1 & 2 - on paper

Not clear about 3 and 4

## Meaningful plots of dependence of binary targets on inputs

1. Come up with a plot that conveys useful information about the dependence (if any) of the target t on the input x.

```{r}
setwd('~/Box Sync/abarciausksas/myfiles/Statistical Modelling and Inference/week4')
data <- read.delim('binary_regression.txt', sep = ' ')

# Visually, they seem uncorrelated
data$tfact <- as.factor(data$t)
plot(data$x, jitter(data$t, factor = 0.5), col = c("red", "blue")[data$tfact])
model <- glm(formula = data$t ~ data$x, family=binomial(link="logit"))
# Not statiscally significant, which matches the visual intuition
summary(model)

intercept <- coefficients(model)[[1]]
p1 <- coefficients(model)[[2]]

library(boot)
inv.logit(intercept + p1*mean(data$x))
# 0.4739566 -> at the mean value of x the prob of t = 1 is 0.48

data$fitted.probs <- inv.logit(intercept + p1*data$x)
plot(data$x, data$fitted.probs)
# A straight line of points indicating probabilities from 0.44 to 0.52
# > summary(data$fitted.probs)
#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.4426  0.4548  0.4694  0.4740  0.4909  0.5315 
# There are more observations at higher values of x - is this having some impact on the linear distribution of the inv.logit function?
```

2. Meaningful plots of dependence of binary targets on inputs

**Download the dataset labelled_suspicious.txt. Come up with a plot that conveys useful information about the dependence (if any) of the target t on the features $\phi1$ and $\phi2$.**

```{r}
setwd('~/Box Sync/abarciausksas/myfiles/Statistical Modelling and Inference/week4')
data <- read.delim('labelled_suspicious.txt', sep = ' ')

phi1model <- glm(formula = data$t ~ data$phi1, family=binomial(link="logit"))
phi2model <- glm(formula = data$t ~ data$phi2, family=binomial(link="logit"))
phimodel <- glm(formula = data$phi1 ~ data$phi2, family=binomial(link="logit"))

summary(phi1model)
summary(phi2model)
summary(phimodel)
plot(data$phi1, data$phi2, ylim = c(-1,2), xlim = c(-1,2))

# This analysis shows evidence that phi1 and phi2 are highly correlated (phi1 is 0.5 of phi2)
```
